Supporting Information:\\Black-swan events in animal populations

### 2016-07-05 text/supp.tex
Copy editing and rearranging of supp for pnas

\subsubsection{Autocorrelated residuals} We considered a version of the Gompertz
model in which an autoregressive parameter was fit to the process-noise
residuals:
\begin{align}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, \phi \epsilon_{t-1}, \sigma).
\end{align}
In addition to the parameters in the original Gompertz model, this model
estimates an additional parameter \(\phi\), which represents the correlation of
subsequent residuals. Based on the results of previous analyses with the
GPDD \cite{connors2014} and the chosen prior in previous
analyses \cite{thorson2014a}, we placed a moderately informative prior on
\(\phi\) that assumed the greatest probability density near zero with the
reduced possibility of \(\phi\) being near \(-1\) or \(1\). Specifically, we
chose \(\phi \sim \mathrm{truncated\mhyphen Normal}(0, 0.5, \mathrm{min.} = -1,
\mathrm{max.} = 1)\).

\subsubsection{Density independence} We fit a simplified version of the
Gompertz model in which the density dependence parameter \(b\) was fixed at
\(1\) (density independent). This is equivalent to fitting a random walk model
(with drift) to the \(\log\) abundances or assuming the growth rates are drawn
from a stationary distribution. The model was as follows:
\begin{align}
x_t &= \lambda + x_{t-1} + \epsilon_t\\
\epsilon &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma).
\end{align}
We fit this model for three reasons: (1) it is computationally simpler and so
provides a check that our more complicated full Gompertz model was obtaining
reasonable estimates of \(\nu\), (2) it provides a test of whether density
dependence was affecting our perception of heavy tails, (3) it
matches how some previous authors have modelled heavy tails without accounting
for density dependence \cite{segura2013}.

\subsubsection{Observation error} Observation error can bias parameter
estimates \cite{knape2012} and is known to affect the ability to detect extreme
events \cite{ward2007}. In our main analysis, we fit a model that ignored
observation error. One way to account for observation error would be to fit a
full state-space model that simultaneously estimates the magnitude of process
noise and observation error. However, simultaneously estimating observation and
process noise is a challenging problem (e.g.\ because the observation and
process noise parameters tend to negatively covary in model fitting) and is
known to sometimes result in identifiability issues with the Gompertz
population model \cite{knape2008}. Furthermore, our model was applied to
hundreds of time series, often of short length (as few as \minTimeSteps\ time
steps) and our model estimates an additional parameter---the shape of the
process deviation tails---potentially making identifiability and computational
issues even greater. Therefore, we considered a version of the base Gompertz
model that allowed for a fixed level of observation error:
\begin{align}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma_\mathrm{proc}),
\end{align}
where \(U\) represents the served state vector, \(\sigma_\mathrm{obs}\)
represents the standard deviation of observation error (on a log scale), and
\(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set
\(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
simulation analyses \cite{valpine2002, thorson2014b}.

\subsubsection{Ricker-logistic} We also fit a Ricker-logistic model:
\begin{align}
x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma),
\end{align}
where \(r_\mathrm{max}\) represents the theoretical maximum population growth
rate that is obtained when \(N_t = 0\). The parameter \(K\) represents the
carrying capacity and, as before, \(x_t\) represents the \(\log\) transformed
abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in
population growth rate with increases in abundance. In contrast, the Gompertz
model assumes a linear decrease in population growth rate with increases in
\textit{log} abundance.

To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between
zero and twice the maximum observed abundance (similar and less informative
than in Ref.~\cite{delean2013}). We set the prior on
\(r_\mathrm{max}\) as uniform between 0 and 20 as in Ref.~\cite{delean2013}. We used the same priors on \(\nu\) and \(\sigma\) as in
the Gompertz model.

\subsection{Alternative priors}

To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
we refit our models with weaker and stronger priors. Our base model used a
prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
used Truncated-Exponential(0.02, min.\ = 2) (Fig.~S\ref{fig:priors}c). Note that
the base and weaker priors are relatively flat within the region of \(\nu <
20\), which is the region we are concerned about when categorizing
populations as heavy tailed. We also considered using a Uniform(\(0, 1\))
prior on \(1/\nu\) as used in Ref.~\cite{gelman2014}, but this
prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
than the truncated exponential priors considered here,
and the results are therefore not shown.

\subsection{Simulation testing}
We performed two types of simulation testing.
First, we tested how easily the Student-$t$ distribution \(\nu\) parameter could
be recovered given different true values of \(\nu\) and different sample sizes.
Second, we tested the ability of the heavy-tailed Gompertz model to obtain
unbiased parameter estimates of \(\nu\) given that a set of process deviations
was provided in which the effective \(\nu\) value was close to the true \(\nu\)
value.

We found that we could consistently and precisely recover median posterior
estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge
200\)) (Fig.~S\ref{fig:sim-nu} upper panels). At smaller samples we could
usually qualitatively distinguish heavy from not-heavy tails, but the model
tended to underestimate how heavy the tails were. At the same time, at smaller
sample sizes, the model tended to overestimate how large the scale parameter
was (Fig.~S\ref{fig:sim-nu} lower panels).

Our simulation-testing results indicate that the Gompertz model can recapture
the true value of \(\nu\) when the process noise was chosen so that appropriate
tail events were present (Fig.~S\ref{fig:sim-prob} upper panels). The addition
of observation error caused the model to tend to underestimate the degree of
heavy-tailedness. Fitting a model with correctly specified observation error
did not substantially reduce model bias (Fig.~S\ref{fig:sim-prob}).

@@

@@

First, we tested the ability to estimate $\nu$ given different true values of
$\nu$ and different sample sizes. We took stochastic draws from t
distributions with different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$
[$\approx$ normal]), with central tendency parameters of $0$, and scale
parameters of $1$. We started with $1600$ stochastic draws and then fit the
models again at the first $800, 400, 200, 100, 50,$ and $25$ draws. Each time
we recorded the posterior samples of $\nu$.

@@

@@

\begin{equation}
  \mathrm{Skewed\,Student}\mhyphen t(x) =
\begin{dcases}
\frac{2}{(\gamma + 1/\gamma)}
         \mathrm{Student}\mhyphen t(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
       \frac{2}{(\gamma + 1/\gamma)}
         \mathrm{Student}\mhyphen t(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
\end{dcases}
\end{equation}

@@

\begin{equation}
\mathrm{Skewed\,Student}\mhyphen t(x) =
\begin{dcases}
  \log (2 \gamma) - \log(\gamma^2 + 1) +
  \log \mathrm{Student}\mhyphen t(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
  \log (2 \gamma) - \log(\gamma^2 + 1) +
  \log \mathrm{Student}\mhyphen t(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
\end{dcases}
\end{equation}

@@

%\clearpage
\section{Additional acknowledgements}

Many of the silhouette images used in Figs 2, and 3 were obtained from
\texttt{phylopic.org} under Creative Commons licenses. We vectorized the
salmon in Fig.~2. The bird in these figures was obtained
from \texttt{phylopic.org} under a Creative Commons Attribution 3.0 Unported
license with credit to Jean-Raphaël Guillaumin {[}photography{]} and T.
Michael Keesey {[}vectorization{]}). The silhouettes in
Fig.~3 were obtained from the following sources (metadata
obtained with the help of the rphylopic R package,
\url{https://github.com/sckott/rphylopic}):

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=1.1\textwidth]{../analysis/heavy-skew-projections-ggplot}
\caption{Projections of population abundance under skew-$t$ (red) and
normally distributed process error models (blue) for populations with
Pr($\nu<10$) $>0.5$. Numbers above panels indicate GPDD population IDs as
shown in table~S2. Horizontal axes indicates year before last
recorded abundance and a projection five years into the future. For the
projected region, solid lines represents the median projection and
dashed lines represent lower 99\% credible intervals. Vertical axes are
log-distributed. Assuming normal process error tends to underestimate risk of
low abundance if the process error is in fact heavy-tailed.}

>>>>

\section{Prior distributions for Gompertz models}

For the Gompertz model, our weakly informative priors (Fig.~\ref{fig:priors}) were:
\begin{align}
b &\sim \mathrm{Uniform}(-1, 2)\\
\lambda &\sim \mathrm{Normal}(0, 10^2)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\
\nu &\sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
\end{align}
Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
would not expect values of \(b\) with levels of density dependence as low as
\(-1\) (very strong inverse density dependence), nor would we generally expect
values above \(1\). We allowed values of \(b\) above \(1\) to allow for
non-stationary time series of growth rates. The estimates of \(b\) were well
within these bounds. Our prior on \(\lambda\) was weakly informative within the
range of expected values for population growth and is similar to the default
priors suggested by \citet{gelman2008d} for intercepts of
regression models. Our half-Cauchy prior on \(\sigma\) follows
\citet{gelman2006c} and \citet{gelman2008d} and the specific scale parameter of
\(2.5\) is based on our expected range of the value in nature from previous
studies \citep{connors2014}. In our testing of a subsample of populations, our
parameter estimates were not qualitatively changed by switching to a uniform
prior on \(\sigma\), but the weakly informative half-Cauchy prior sped up chain
convergence.

To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
we refit our models with weaker and stronger priors. Our base model used a
prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
used Truncated-Exponential(0.02, min.\ = 2) (Fig.~\ref{fig:priors}c). Note that
the base and weaker priors are relatively flat within the region of \(\nu <
20\), which is the region we are concerned about when categorizing
populations as heavy tailed. We also considered using a Uniform(\(0, 1\))
prior on \(1/\nu\) as used in \citet{gelman2014}, but this
prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
than the truncated exponential priors considered here,
and the results are therefore not shown.

\section{Alternative population models}

\subsection{Autocorrelated residuals} We considered a version of the Gompertz
model in which an autoregressive parameter was fit to the process-noise
residuals:
\begin{align}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, \phi \epsilon_{t-1}, \sigma).
\end{align}
In addition to the parameters in the original Gompertz model, this model
estimates an additional parameter \(\phi\), which represents the correlation of
subsequent residuals. Based on the results of previous analyses with the
GPDD \citep{connors2014} and the chosen prior in previous
analyses \citep{thorson2014a}, we placed a moderately informative prior on
\(\phi\) that assumed the greatest probability density near zero with the
reduced possibility of \(\phi\) being near \(-1\) or \(1\). Specifically, we
chose \(\phi \sim \mathrm{truncated\mhyphen Normal}(0, 0.5, \mathrm{min.} = -1,
\mathrm{max.} = 1)\).

\subsection{Density independence} We fit a simplified version of the
Gompertz model in which the density dependence parameter \(b\) was fixed at
\(1\) (density independent). This is equivalent to fitting a random walk model
(with drift) to the \(\log\) abundances or assuming the growth rates are drawn
from a stationary distribution. The model was as follows:
\begin{align}
x_t &= \lambda + x_{t-1} + \epsilon_t\\
\epsilon &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma).
\end{align}
We fit this model for three reasons: (1) it is computationally simpler and so
provides a check that our more complicated full Gompertz model was obtaining
reasonable estimates of \(\nu\), (2) it provides a test of whether density
dependence was affecting our perception of heavy tails, (3) it
matches how some previous authors have modelled heavy tails without accounting
for density dependence \citep{segura2013}.

\subsection{Observation error} Observation error can bias parameter
estimates \citep{knape2012} and is known to affect the ability to detect extreme
events \citep{ward2007}. In our main analysis, we fit a model that ignored
observation error. One way to account for observation error would be to fit a
full state-space model that simultaneously estimates the magnitude of process
noise and observation error. However, simultaneously estimating observation and
process noise is a challenging problem (e.g.\ because the observation and
process noise parameters tend to negatively covary in model fitting) and is
known to sometimes result in identifiability issues with the Gompertz
population model \citep{knape2008}. Furthermore, our model was applied to
hundreds of time series, often of short length (as few as \minTimeSteps\ time
steps) and our model estimates an additional parameter---the shape of the
process deviation tails---potentially making identifiability and computational
issues even greater. Therefore, we considered a version of the base Gompertz
model that allowed for a fixed level of observation error:
\begin{align}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma_\mathrm{proc}),
\end{align}
where \(U\) represents the served state vector, \(\sigma_\mathrm{obs}\)
represents the standard deviation of observation error (on a log scale), and
\(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set
\(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
simulation analyses \citep{valpine2002, thorson2014b}.

\subsection{Ricker-logistic} We also fit a Ricker-logistic model:
\begin{align}
x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma),
\end{align}
where \(r_\mathrm{max}\) represents the theoretical maximum population growth
rate that is obtained when \(N_t = 0\). The parameter \(K\) represents the
carrying capacity and, as before, \(x_t\) represents the \(\log\) transformed
abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in
population growth rate with increases in abundance. In contrast, the Gompertz
model assumes a linear decrease in population growth rate with increases in
\textit{log} abundance.

To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between
zero and twice the maximum observed abundance (similar and less informative
than in \citet{delean2013}). We set the prior on
\(r_\mathrm{max}\) as uniform between 0 and 20 as in \citet{delean2013}. 
We used the same priors on \(\nu\) and \(\sigma\) as in
the Gompertz model.

@@

We performed two types of simulation testing.
First, we tested how easily the Student-$t$ distribution \(\nu\) parameter could
be recovered given different true values of \(\nu\) and different sample sizes.
Second, we tested the ability of the heavy-tailed Gompertz model to obtain
unbiased parameter estimates of \(\nu\) given that a set of process deviations
was provided in which the effective \(\nu\) value was close to the true \(\nu\)
value.

@@

First, we tested the ability to estimate $\nu$ given different true values of
$\nu$ and different sample sizes. We took stochastic draws from $t$
distributions with different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$
[$\approx$ normal]), with central tendency parameters of $0$, and scale
parameters of $1$. We started with $1600$ stochastic draws and then fit the
models again at the first $800, 400, 200, 100, 50,$ and $25$ draws. Each time
we recorded the posterior samples of $\nu$.

We found that we could consistently and precisely recover median posterior
estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge
200\)) (Fig.~\ref{fig:sim-nu} upper panels). At smaller samples we could
usually qualitatively distinguish heavy from not-heavy tails, but the model
tended to underestimate how heavy the tails were. At the same time, at smaller
sample sizes, the model tended to overestimate how large the scale parameter
was (Fig.~\ref{fig:sim-nu} lower panels).

@@

We found that the Gompertz model can recapture
the true value of \(\nu\) when the process noise was chosen so that appropriate
tail events were present (Fig.~\ref{fig:sim-prob} upper panels). The addition
of observation error caused the model to tend to underestimate the degree of
heavy-tailedness. Fitting a model with correctly specified observation error
did not substantially reduce model bias (Fig.~\ref{fig:sim-prob} upper panels).

@@

\begin{equation}
  \mathrm{Skewed\,Student}\mhyphen t(x) =
\begin{dcases}
\frac{2}{(\gamma + 1/\gamma)}
         \mathrm{Student}\mhyphen t(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
       \frac{2}{(\gamma + 1/\gamma)}
         \mathrm{Student}\mhyphen t(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0.
\end{dcases}
\end{equation}

@@

\begin{equation}
\mathrm{Skewed\,Student}\mhyphen t(x) =
\begin{dcases}
  \log (2 \gamma) - \log(\gamma^2 + 1) +
  \log \mathrm{Student}\mhyphen t(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
  \log (2 \gamma) - \log(\gamma^2 + 1) +
  \log \mathrm{Student}\mhyphen t(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0.
\end{dcases}
\end{equation}

@@

%\clearpage
\section{Silhouette image licenses}

Many of the silhouette images used in Figs 2, and 3 were obtained from
\texttt{phylopic.org} under Creative Commons licenses. We drew the
salmon in Fig.~2. The bird in these figures was obtained
from \texttt{phylopic.org} under a Creative Commons Attribution 3.0 Unported
license with credit to Jean-Raphaël Guillaumin {[}photography{]} and T.
Michael Keesey {[}vectorization{]}). The silhouettes in
Fig.~3 were obtained from the following sources (metadata
obtained with the help of the rphylopic R package,
\url{https://github.com/sckott/rphylopic}):

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=1.1\textwidth]{../analysis/heavy-skew-projections-ggplot}
\caption{Projections of population abundance under skew-$t$ (red) and
normally distributed process error models (blue) for populations with
Pr($\nu<10$) $>0.5$. Numbers above panels indicate GPDD population IDs as
shown in Table~S2. Horizontal axes indicates year before last
recorded abundance and a projection five years into the future. For the
projected region, solid lines represents the median projection and
dashed lines represent lower 99\% credible intervals. Vertical axes are
log-distributed. Assuming normal process error tends to underestimate risk of
low abundance if the process error is in fact heavy-tailed.}
